Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-04-19 VegOzone.tex
Restructure Methodes and results

In Section~\ref{sec:dam}, we give a brief summary of the key components and  drivers of the $\mathrm{DO_3SE}$ model that is used to assess the ozone risk. To study the influence of selected parameters on ozone risk in the subarctic, we propose a method to localize temperature response functions based on climate data and start and end of the growing season from satellite observed gross primary productivity (GPP). For the later discussion of results, we summarize climate and weather conditions of the 2018/19 growing season at our research site in Section~\ref{sec:stats}. Results of the localization and ozone risk modelling are presented in Section~\ref{sec:results}. In Section~\ref{sec:conc}, we discuss these results concerning prevailing conditions and give an outlook on how future assessments of pollution risk in the subarctic could be improved.

@@

To relate local ozone concentration \chem{[O_3]} with visible damage on ozone sensitive plants, an ozone bio-monitor (referred to as ozone garden) has been installed at Svanhovd in the 2018/19 GS. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3).

@@

We deduced the dominant species on-site at Svanhovd from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations in the following. For coniferous trees, we base our localization of parameters on MM's Boreal Norway spruce. Our deciduous tree paramterization is based on MM's silver birch and perennial grassland on MM's perennial grassland for Central Europe. For a comprehensive list of generic MM model parameters, consult Supplement~Table~S1.

\subsubsection{Temperature acclimation}

Initial test simulations with the MM PFTs revealed an unrealistically low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance in this species occurred only during an extended warm period in late July. In ecological terms, this means that there was almost no growth of grass in the summer of 2019 -- a prediction that is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low throughout the season, being the limiting factor of stomatal conductance in perennial grassland. Note, these parameterizations are hypothetical and have yet to be verified by experiments.

We define two different hypothetical temperature acclimations: subarctic and cold. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. This is accomplished by moving $f_\mathrm{T}$ towards cooler temperatures but keep $T_\mathrm{max}$ at its MM value. In the same way, \emph{subarctic} is constructed to represent a species that is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shift $f\mathrm{T}$ to even colder temperatures and choose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. In other words, we aim to increase the overlapping area between $f_\mathrm{T}$ and the temperature probability density function (PDF) during the GS. The most extreme of the two acclimation types is \emph{subarctic}.
In Fig.~\ref{fig:f_temp_grassland}a, we show the resulting $f_\mathrm{T}$ for perennial grassland together with temperature PDFs deduced from the five months GS in two climatological periods. The periods 1990s and 2000s have been chosen to indicate that the vegetation has been subject to climate change already. Thus, the predominant acclimation at present is not known.

The resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}a and \ref{fig:f_temp_birch}a, respectively. We base our temperature acclimation of coniferous trees on experimental results on Norway spruce which were found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. In particular, we use the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland by \citet{TB:Kolari2007} to estimate the optimal temperature interval ($10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$). We presume a similar temperature acclimation for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.

\subsubsection{Light acclimation}
TODO:
Regarding the acclimation of $f_\mathrm{light}$, we presume that the opening extent of stomata at low light conditions differs in subarctic species compared to species in less extreme climates. We, therefore, adjust the light sensitivity of stomata such that the PPFD value needed to reach $50\,\unit{\%}$ opening is varied $\pm 20\,\unit{\%}$. For this we derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) analytically for each PFT or species $k$
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value $\alpha_\mathrm{MM}^k$ at $0.5$ 
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. By inverting Eq.~(\ref{eq:inverse_function}) for $\alpha$, we find $\alpha'(\gamma')$. The resulting functions for perennial grassland (PPFD0.8, PPFD1.2) are shown in Fig.~\ref{fig:f_temp_grassland}b) (for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b and \ref{fig:f_temp_birch}b. All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.

@@

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{javis_func_opt_mean}
  %\includegraphics[width=8.3cm]{fig08}
  \caption{Proposed metric to test bespoke response functions. GS (May--August) climatological averages and standard deviation of $g_\mathrm{sto}^k$ (Eq.~(\ref{eq:stomatal})) relative to $g_\mathrm{max}^k$ at noon (averaged over $11\,\unit{am}-1\,\unit{pm}$ local time) and in the morning (averaged over $5-9\,\unit{am}$ local time) are shown for (a) deciduous trees; (b) coniferous trees; (c) perennial grassland. }
  \label{fig:javis_func_opt_mean}
\end{figure}

\subsubsection{Growing season estimate}

Another important factor is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. For coniferous trees, we use a MODIS (Aqua/Terra) retrieved net photosynthesis product \citep{MODIS_PSN} for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis vs. time of year data can be approximated by a second-order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start}$ as \unit{doy}~122 for 2018 and \unit{doy}~106 in 2019. $A_\text{end}$ amounts to \unit{doy}~261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 is, hence, one month longer than in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}). We find \unit{doy}~129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR). For comparison, the generic dates for central Europe ($A_\mathrm{start} = 100$ and $A_\mathrm{end} = 307$) indicating a substantially longer growing season. Due to a lack of quantitative field observation at Svanhovd, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}.

@@

\section{Climate and weather conditions}
\label{sec:stats}
Here, we put the conditions during the growing season 2018/19 into a climatological context.

\subsection{Weather conditions in the growing season 2018/19}
\label{subsec:weather}

@@

\subsection{Climate (1990-present)}
\label{subsec:climatologies}

>>>>

In Section~\ref{sec:dam}, we give a brief summary of the key components and  drivers of the $\mathrm{DO_3SE}$ model that is used to assess the ozone risk. To study the influence of selected parameters on ozone risk in the subarctic, we propose a method to localize temperature response functions based on climate data and start and end of the growing season from satellite observed gross primary productivity (GPP). For the later discussion of results, we summarize climate and weather conditions of the 2018/19 growing season at our research site in Section~\ref{sec:stats}. Results of the localization, the ozone risk modelling, and the climate perspective are presented in Section~\ref{sec:results}. In Section~\ref{sec:conc}, we discuss these results concerning prevailing conditions and give an outlook on how future assessments of pollution risk in the subarctic could be improved.

@@

To relate local ozone concentration \chem{[O_3]} with visible damage on ozone sensitive plants, an ozone bio-monitor (referred to as ozone garden) has been installed at Svanhovd in the 2018/19 GS. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3). Note that the extent of visible damage is not equivalent to a quantitative reduction of above/below ground biomass of a species.

@@

We deduced the dominant species on-site at Svanhovd and from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations. The localization of parameters for coniferous trees is based on MM's Boreal Norway spruce, deciduous trees on MM's silver birch, and perennial grassland on MM's perennial grassland for Central Europe. For a comprehensive list of MM parameters, consult Supplement~Table~S1. In the following, we suggest localizations of key environmental parameters to a subarctic climate based on meteorological data to assess over-/underestimation of ozone damage by using MM Boreal parameterizations.

\subsubsection{Temperature and light response}

Initial simulations with the MM parameterizations for perennial grassland revealed an unexpected low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance occurred only during an extended warm period in late July. In ecological terms, this means that there would have been almost no growth of grass in the summer of 2019 -- a model prediction that is falsified by reality. We identified $f_\mathrm{T}$ as abnormally low throughout the season, being the limiting factor of stomatal conductance in perennial grassland.

To localize the temperature response function $f_\mathrm{T}$, we divide the temperature dataset into two subsets: 1992--2000 (near past) and 2011--2019 (present). From these two subsets, we select dates within the GS (May--September) only and plot the probability density function (PDF) as histogram with $1\,\unit{^\circ C}$ binning. We overlay the histograms with $f_\mathrm{T}$ to graphically determine the goodness of fit of the MM parameterization from the overlapping area. In Fig.~\ref{fig:f_temp_grassland}a, we show the $f_\mathrm{T}$ for perennial grassland together with the temperature PDFs for near past and present climate conditions. The MM parameterization is displayed as a solid line. A comparison between the two PDFs indicates that the temperature distribution has been shifted towards higher temperatures reducing the number of days below $5\,\unit{^\circ C}$. It is unclear how fast the species composition of perennial grassland responds to these changes. Thus, the predominant acclimation at present is not known.

The idea is to define a localized temperature response function by increasing the overlapping area between $f_\mathrm{T}$ and the temperature PDF during the GS. To this end, we define two different hypothetical target species: subarctic and cold. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures than MM, but slightly less efficient at warm temperatures than MM. This is accomplished by moving $T_\mathrm{min}$ and $T_\mathrm{opt}$ towards cooler temperatures but keep $T_\mathrm{max}$ at its MM value. In the same way, \emph{subarctic} is constructed to represent a species that is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shift both $T\mathrm{min}$ and $T_\mathrm{max}$ to colder temperatures and choose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. The most extreme of the two localizations compared to MM is \emph{subarctic}.

For localizing $f_\mathrm{light}$, we presume that the opening extent of stomata at low light conditions differs in subarctic species compared to species in less extreme climates. We, therefore, adjust the modelled light sensitivity of stomata such that the PPFD value needed to reach $50\,\unit{\%}$ opening is varied by $\pm 20\,\unit{\%}$. For this, we derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) for each PFT $k$ analytically
%
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
%
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value of $\alpha_\mathrm{MM}^k$ at $0.5$ ($50\,\%$ opening)
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. By solving Eq.~(\ref{eq:inverse_function}) for $\alpha$, we find $\alpha'(\gamma')$.

In Fig.~\ref{fig:f_temp_grassland}b, we show the PPFD PDF for near past and present together with the $f_\mathrm{light}$ for perennial grassland. The MM parameterization is displayed as solid line.

@@

Similarly, we defined localized $f_\mathrm{T}$ and $f_\mathrm{light}$ for coniferous and deciduous trees. The resulting response functions are shown in Appendix~Figs.~\ref{fig:f_temp_spruce} and \ref{fig:f_temp_birch}, respectively.

Coniferous trees are able to start photosynthesis as soon as the leaf temperature reaches a certain threshold \citep{TB:Kolari2007} and are, hence, found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 but cease activity in less favorable conditions \citep{TB:Kolari2007, TP:Wallin2013}.

To determine the temperature localization for coniferous trees, we use the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland by \citet{TB:Kolari2007}. We estimate an optimal temperature interval ($10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$). For simplicity, we assume the same interval for deciduous trees.

The resulting parameterizations are presented in detail in Section~\ref{subsec:res_local}.

\subsubsection{Growing season estimate}
\label{subsec:gs_est}

Another important factor for the accumulation of ozone dose is the timing and length of the GS.

TODO: estimated as follows. For coniferous trees, we use a MODIS (Aqua/Terra) retrieved net photosynthesis product \citep{MODIS_PSN} for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis vs. time of year data can be approximated by a second-order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start}$ as \unit{doy}~122 for 2018 and \unit{doy}~106 in 2019. $A_\text{end}$ amounts to \unit{doy}~261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 is, hence, one month longer than in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}). We find \unit{doy}~129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR). For comparison, the generic dates for central Europe ($A_\mathrm{start} = 100$ and $A_\mathrm{end} = 307$) indicating a substantially longer growing season. Due to a lack of quantitative field observation at Svanhovd, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}.

\subsubsection{Soil conditions}
\label{subsec:soil}

All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.

(for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b and \ref{fig:f_temp_birch}b)

All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{javis_func_opt_mean}
  %\includegraphics[width=8.3cm]{fig08}
  \caption{Proposed metric to test bespoke response functions. GS (May--August) climatological averages and standard deviation of $g_\mathrm{sto}^k$ (Eq.~(\ref{eq:stomatal})) relative to $g_\mathrm{max}^k$ at noon (averaged over $11\,\unit{am}-1\,\unit{pm}$ local time) and in the morning (averaged over $5-9\,\unit{am}$ local time) are shown for (a) deciduous trees; (b) coniferous trees; (c) perennial grassland.}
  \label{fig:javis_func_opt_mean}
\end{figure}

Start ($A_\text{start}$) and end ($A_\text{end}$) of the GS for each PFT are usually given in \unit{doy} (Table~\ref{tab:sensitivity_tests_gs})

@@

\subsection{Climate and weather conditions}
\label{sec:stats}
Here, we put the conditions during the growing season 2018/19 into a climatological context.

\subsubsection{Weather conditions in the growing season 2018/19}
\label{subsec:weather}

@@

\subsubsection{Climate (1990-present)}
\label{subsec:climatologies}
